\section{Conclusions and Future Work}
\label{sec:cafw}

This work presented \textit{rzstd}, a safety-oriented implementation of the Zstandard decompression
algorithm that models entropy decoding as a bounded finite-state reconstruction process.
By design, the implementation makes state transitions, normalization constraints, and window
bounds explicit within the type system and execution model.

Entropy coding schemes such as Finite State Entropy can be modelled in practice as deterministic
transitions over discrete probability distributions. Therefore, correct decoding depends not
only on reconstruction symbol sequence, but also on preserving structural invariants
throughout evaluation. By encoding these constraints directly into the decoding architecture,
\textit{rzstd} demonstrates how theoretical structures can be mapped to a memory-safe systems
implementation without sacrificing asymptotic efficiency.

Empirical evaluation shows that the decoder achieves throughput comparable to other safe Rust
implementations, though it remains below the heavily optimized C reference library. While performance
differences persist, the results suggest that explicit invariant enforcement and spatial
memory safety can coexist with competitive decompression throughput in user-facing
contexts.

\paragraph{Future Work}

Several directions remain to improve correctness coverage, implementation maturity, and
feature completeness:

\begin{itemize}
  \item \textbf{Correctness and Conformance:} The XXH64 checksum mismatch observed in the
    \texttt{mozilla} corpus entry must be resolved to ensure full conformance with RFC 8878.
    Implementing fuzzing will further evaluate robustness against malformed and adversarial
    inputs.

  \item \textbf{Further Optimization:} Future iterations will focus on streaming sequence
    execution, improving cache locality, and enabling bounds-check elision through
    statically provable invariants to reduce the throughput overhead while preserving
    safety guarantees.

  \item \textbf{Protocol Extensions:} Supporting dictionary-based decompression and the
    Zstandard Seekable Format will broaden the functional scope of the current
    implementation.
\end{itemize}

To summarize, \textit{rzstd} illustrates that entropy-coded signal reconstruction can be expressed
as a deterministic process operating within an explicitly bounded context in a safe systems
language. The work highlights how careful architectural design, memory safety mechanisms and
rigorous alignment with the model can reduce implementation risk while retaining practical performance
characteristics.